Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization \thanks{Identify applicable funding agency here. If none, delete this.}

### 2020-07-16 latex/tex/introduction.tex

Hierarchical clustering is the process of recursively breaking a group into two similar groups until every sample is a cluster by itself. That is., we consider all the data points as a single cluster and in each iteration, the data points in the current cluster are split into two separate clusters till there is only one sample in a cluster.
While standard NMF is interpreted as a flat clustering, it can also be extended for hierarchical clustering.
Kuang and Park \cite{KP13} propose a method that uses rank-2 NMF to recursively bipartition the samples.
An NMF with $k=2$ determines a row of $\M{H}$ of length 2 for each sample that determines its part membership.
By recursively bipartitioning the samples in each part, the method determines a binary tree such that each sample belongs to a unique leaf and internal nodes of the tree determine higher-level clusters of the samples.
A single latent feature vector for each node can also be used for interpreting clusters.
Similarly, Du et al. \cite{DKDP17} show that the latent features vectors of the leaves can also be used to initialize a $\M{W}$ within an algorithm for flat NMF, yielding fast convergence and overall speedup compared to initializing the algorithm with random data.
We discuss the hierarchical method in more detail in \Cref{sec:prelim} and \Cref{sec:seqalgs}.

We illustrate the output of the hierarchical clustering method with an example data set and output tree.
Following Gillis et al. \cite{GKP15}, we apply the method to a hyperspectral imaging (HSI) data set of the Washington, D.C~national mall, which has pixel dimensions $1280 \times 307$ and 191 spectral bands.
\Cref{fig:dc} visualizes the output tree with 6 leaves along with their hierarchical relationships.
The root node, labeled 0, is a flattening of the HSI data to a 2D grayscale image.
Each other node is represented by an overlay of the member pixels of the clusters (in blue) on the original grayscale image.
The first bipartitioning separates vegetation (cluster 1) from non-vegetation (cluster 2), the bipartitioning of cluster 1 separates grass (cluster 3) from trees (cluster 4), the bipartitioning of cluster 2 separates buildings (cluster 5) from sidewalks/water (cluster 6), and so on.
If the algorithm continues, it chooses to split the leaf node that provides the greatest benefit to the overall tree, which can be quantified as a node's ``score'' in various ways.
The algorithm proceeds to split nodes until a target number of leaves is obtained, until no node is worth splitting, or some other stopping criterion is met.

>>>>

Hierarchical clustering is the process of recursively breaking a group into two similar groups until every sample is a cluster by itself. That is., we consider all the data points as a single cluster and in each iteration, the data points in the current cluster are split into two separate clusters till there is only one sample in a cluster.
While standard NMF is interpreted as a flat clustering, it can also be extended for hierarchical clustering.
Kuang and Park \cite{KP13} propose a method that uses rank-2 NMF to recursively bipartition the samples.
An NMF with $k=2$ determines a row of $\M{H}$ of length 2 for each sample that determines its part membership.
\LM{delete previous sentence?}
By recursively bipartitioning the samples in each part, the method determines a binary tree such that each sample belongs to a unique leaf and internal nodes of the tree determine higher-level clusters of the samples.
\LM{By recursively bipartioning the samples in each node, the method determines a binary tree such that all leaves contain unique samples and the structure of the tree determines hierarchical clusters}
A single latent feature vector for each node can also be used for interpreting clusters.
\LM{A single W vector for each node can also be used for cluster interpretation.}
Similarly, Du et al. \cite{DKDP17} show that the latent features vectors of the leaves can also be used to initialize a $\M{W}$ within an algorithm for flat NMF, yielding fast convergence and overall speedup compared to initializing the algorithm with random data.
We discuss the hierarchical method in more detail in \Cref{sec:prelim} and \Cref{sec:seqalgs}.
\LM{Get rid of this entirely since it isn't really discussed further in 2 and 3?}

We illustrate the output of the hierarchical clustering method with an example data set and output tree.
Following Gillis et al. \cite{GKP15}, we apply the method to a hyperspectral imaging (HSI) data set of the Washington, D.C~national mall, which has pixel dimensions $1280 \times 307$ and 191 spectral bands.
\Cref{fig:dc} visualizes the output tree with 6 leaves along with their hierarchical relationships.
The root node, labeled 0, is a flattening of the HSI data to a 2D grayscale image.
Each other node is represented by an overlay of the member pixels of the clusters (in blue) on the original grayscale image.
The first bipartitioning separates vegetation (cluster 1) from non-vegetation (cluster 2), the bipartitioning of cluster 1 separates grass (cluster 3) from trees (cluster 4), the bipartitioning of cluster 2 separates buildings (cluster 5) from sidewalks/water (cluster 6), and so on.
If the algorithm continues, it chooses to split the leaf node that provides the greatest benefit to the overall tree, which can be quantified as a node's ``score'' in various ways.
The algorithm proceeds to split nodes until a target number of leaves is obtained, until no node is worth splitting, or some other stopping criterion is met.
\LM{Should we have text explaining it or could we add these labels to the figure itself?}
